# (PART) DATA PREPROCESSING {-}

# Data Tidying {#data-tidying}

## Prerequisite
- Output from the bioinformatics pipelines such as:
 - Shared and conserved taxonomy tables from `mothur` pipeline.
 - Feature and taxonomy tables from `qiime2` pipeline.
 - Microbial profiles files from `metaphlan` pipeline.

## Import libraries and data
```{r}
set.seed(110912)
source("R/common.R")
library(tidyverse, suppressPackageStartupMessages())

cat("Final bioinformatics objects\n\n")
load("../iMAP-part2/RDataRDS/final-bioinfo_objects.RData", verbose = T)
```


## Creating tidy `mothur` otu and taxonomy tables {#mothur-otu-taxonomy}
> Using phylotype data

```{r}
library(tidyverse)
mo_otutable_long <- mo_phylo_shared_table %>% 
  # select(-label, -numOtus) %>% 
  select(-label, -numPhylos) %>% 
  rename(sample_id = Group) %>% 
  rename_all(tolower) %>% 
  pivot_longer(-sample_id, names_to="otu", values_to="count") %>% 
  filter(count != 0)
saveRDS(mo_otutable_long, "RDataRDS/mo_otutable_long.rds")

mo_otutable <- mo_otutable_long %>%
  pivot_wider(id_cols = otu, names_from = sample_id, values_from = count) %>%
  mutate_if(is.numeric, ~replace(., is.na(.), 0))

cat("Dimension of the otutable\n")
dim(mo_otutable)
cat("\nStructure of the otutable\n")
head(mo_otutable)[, 1:3] %>% as.data.frame()  
saveRDS(mo_otutable, "RDataRDS/mo_otutable.rds")


taxlevels <- c("kingdom", "phylum", "class", "order", "family", "genus")
mo_taxonomy <- mo_phylo_cons_taxonomy %>%
  rename_all(tolower) %>%
  select(otu, taxonomy) %>%
  mutate(otu = tolower(otu),
         taxonomy = str_replace_all(taxonomy, "\\(\\d+\\)", ""),
         taxonomy = str_replace(taxonomy, ";unclassified", "_unclassified"),
         taxonomy = str_replace_all(taxonomy, ";unclassified", ""),
         taxonomy = str_replace_all(taxonomy, ";$", "")) %>% 
  separate(taxonomy, into = all_of(taxlevels), sep = ";")

cat("Dimension of the taxonomy table\n")
dim(mo_taxonomy)
cat("\nStructure of the taxonomy table\n")
head(mo_taxonomy)[, 1:4] %>% as.data.frame() 
saveRDS(mo_taxonomy, "RDataRDS/mo_taxonomy.rds")

save(mo_otutable, mo_otutable_long, mo_taxonomy, file = "RDataRDS/otutaxonomy_objects.RData")
save(mo_otutable, mo_otutable_long, mo_taxonomy, file = "RDataRDS/saved_objects.RData")
```


## Creating tidy `qiime2` taxa abundance table {#qiime2-otu-taxonomy}
- Here we read through the merged feature-taxonomy-abundance table and extract abundance at a taxon lineage.
- The demo below extract and split abundance at genus-lineage.
- Additional processing is also done to create a tidy dataframe.
- The final out put will contain feature ID, taxonomy and abundance for each sample.
```{r}
taxlevels <- c("kingdom", "phylum", "class", "order", "family", "genus")

q2_taxa_abund <- q2_feature_taxonomy %>%
  relocate(Taxon, .after = id) %>% 
  mutate_at(3:ncol(.), as.numeric) %>% 
  rename(feature = id,
         taxonomy = Taxon) %>% 
  relocate(taxonomy, .after = feature) %>% 
  distinct() %>%
  mutate_if(is.numeric, ~replace(., is.na(.), 0)) %>%
  mutate(taxonomy = str_replace_all(taxonomy, "; s__$", ""),
         taxonomy = str_replace_all(taxonomy, "; g__$", ""),
         taxonomy = str_replace_all(taxonomy, "; f__$", ""),
         taxonomy = str_replace_all(taxonomy, "; o__$", ""),
         taxonomy = str_replace_all(taxonomy, "; c__$", ""),
         taxonomy = str_replace_all(taxonomy, "; p__$", ""),
         taxonomy = str_replace_all(taxonomy, "; k__$", ""),
         taxonomy = str_replace_all(taxonomy, "\\[|\\]", ""),
         taxonomy = str_replace_all(taxonomy, "\\s", "")) %>%
  dplyr::filter(!grepl("s__*", taxonomy)) %>%
  dplyr::filter(grepl("g__*", taxonomy)) %>% 
  select(-Confidence) %>% 
  mutate(taxonomy = str_replace_all(taxonomy, "\\w__", "")) %>% 
  separate(taxonomy, into = all_of(taxlevels), sep = ";")

colnames(q2_taxa_abund)
saveRDS(q2_taxa_abund, "RDataRDS/q2_taxa_abund.rds")


taxlevels <- c("kingdom", "phylum", "class", "order", "family", "genus")

q2_otutable <- q2_taxa_abund %>%
  select(-all_of(taxlevels))

cat("Dimension of the otutable\n")
dim(q2_otutable)
cat("\nStructure of the otutable\n")
head(q2_otutable)[, 1:3] %>% as.data.frame()  
saveRDS(q2_otutable, "RDataRDS/q2_otutable.rds")

q2_taxonomy <- q2_taxa_abund %>%
  rename_all(tolower) %>%
  select(feature, all_of(taxlevels)) %>%
  distinct()

cat("Dimension of the taxonomy table\n")
dim(q2_taxonomy)
cat("\nStructure of the taxonomy table\n")
head(q2_taxonomy)[, 1:3] %>% as.data.frame() 
saveRDS(q2_taxonomy, "RDataRDS/q2_taxonomy.rds")

resave(q2_otutable, q2_taxonomy, file = "RDataRDS/otutaxonomy_objects.RData")
resave(q2_otutable, q2_taxonomy, file = "RDataRDS/saved_objects.RData")
```


## Creating tidy `metaphlan` relative abundance and taxonomy tables {#m3-otu-taxonomy}
- Here we read through the merged abundance table and extract abundance at a taxon lineage.
- The demo below extract and split abundance at genus-lineage and add code each row to a unique OTU ID.
- Additional processing is also done to create a tidy dataframe.
- The final out put will contain OTU ID, taxonomy and abundance for each sample.

### Standard merged relative abundance
```{r}
taxlevels <- c("kingdom", "phylum", "class", "order", "family", "genus")

m3_taxa_abund <- m3_merged_abund_table %>% 
  select(sample_id, taxonomy, rel_abund) %>% 
  distinct() %>% 
  dplyr::filter(!grepl("\\|s__*", taxonomy),
                grepl("g__*", taxonomy)) %>%
  group_by(sample_id, taxonomy, rel_abund) %>%
  summarise(count = sum(rel_abund),
            rel_abund = rel_abund/100, .groups = "drop") %>% 
  pivot_wider(id_cols = "taxonomy", names_from = "sample_id", values_from = "rel_abund") %>% 
  mutate(otu = paste("otu", seq(1:length(rownames(.))), sep = ""), .before = 1) %>% 
  mutate_if(is.numeric, ~replace(., is.na(.), 0 )) %>% 
  mutate(taxonomy = str_replace_all(taxonomy, "\\w__", "")) %>%
  separate(taxonomy, into = all_of(taxlevels), sep = "\\|") 

saveRDS(m3_taxa_abund, "RDataRDS/m3_taxa_abund.rds")
resave(m3_taxa_abund, file = "RDataRDS/saved_objects.RData")

m3_otutable <- m3_taxa_abund %>% 
  select(-all_of(taxlevels))

cat("Dimension of the otutable\n")
dim(m3_otutable)
cat("\nStructure of the otutable\n")
head(m3_otutable)[, 1:5] %>% as.data.frame()  
saveRDS(m3_otutable, "RDataRDS/m3_otutable.rds")

m3_taxonomy <- readRDS("RDataRDS/m3_taxa_abund.rds") %>% 
  select(otu, all_of(taxlevels))

cat("Dimension of the taxonomy table\n")
dim(m3_taxonomy)
cat("\nStructure of the taxonomy table\n")
head(m3_taxonomy)[, 1:4] %>% as.data.frame() 
saveRDS(m3_taxonomy, "RDataRDS/m3_taxonomy.rds")

resave(m3_otutable, m3_taxonomy, file = "RDataRDS/otutaxonomy_objects.RData")
resave(m3_otutable, m3_taxonomy, file = "RDataRDS/saved_objects.RData")
```


### Relative abundance based on read count
- Here we read through the merged abundance table and extract abundance at a taxon lineage.
- The demo below extract and split abundance at genus-lineage and add code each row to a unique OTU ID.
- Additional processing is also done to create a tidy dataframe.
- The final out put will contain OTU ID, taxonomy and abundance for each sample.

```{r}
taxlevels <- c("kingdom", "phylum", "class", "order", "family", "genus")

m3_read_count_taxa_abund <- m3_merged_abund_table %>% 
  select(sample_id, taxonomy, mapped_read) %>% 
  distinct() %>% 
  dplyr::filter(!grepl("\\|s__*", taxonomy),
                grepl("g__*", taxonomy)) %>%
  pivot_wider(id_cols = "taxonomy", names_from = "sample_id", values_from = "mapped_read") %>% 
  mutate(otu = paste("otu", seq(1:length(rownames(.))), sep = ""), .before = 1) %>% 
  mutate_if(is.numeric, ~replace(., is.na(.), 0 )) %>% 
  mutate(taxonomy = str_replace_all(taxonomy, "\\w__", "")) %>%
  separate(taxonomy, into = all_of(taxlevels), sep = "\\|") 

m3_read_counts_otutable <- m3_read_count_taxa_abund %>% 
  select(-all_of(taxlevels))

cat("Dimension of the otutable\n")
dim(m3_read_counts_otutable)
cat("\nStructure of the otutable\n")
head(m3_read_counts_otutable)[, 1:5] %>% as.data.frame()  
saveRDS(m3_read_counts_otutable, "RDataRDS/m3_read_counts_otutable.rds")

m3_read_counts_taxonomy <- readRDS("RDataRDS/m3_taxa_abund.rds") %>% 
  select(otu, all_of(taxlevels))

cat("Dimension of the taxonomy table\n")
dim(m3_read_counts_taxonomy)
cat("\nStructure of the taxonomy table\n")
head(m3_read_counts_taxonomy)[, 1:4] %>% as.data.frame() 
saveRDS(m3_read_counts_taxonomy, "RDataRDS/m3_read_counts_taxonomy.rds")

resave(m3_read_counts_otutable, m3_read_counts_taxonomy, file = "RDataRDS/otutaxonomy_objects.RData")
resave(m3_read_counts_otutable, m3_read_counts_taxonomy, file = "RDataRDS/saved_objects.RData")
```




