# Data Joining

## Creating joined R objects {#r-objects}

```{block, type="tmbinfo", echo=T}
> Creating an object that joins the metadata, microbial abundance, and taxonomy data simplifies the downstream analyses. The composite objects are in long-format and enables us to do any type of analysis we desire when exploring the results associated with **16S rRNA gene** or **metagenomics** sequencing data.
```

Let' get started


### Prerequisite
- Tidy metadata.
- Tidy microbial abundance table.
- Tidy taxonomy table.

### Load libraries and data
```{r message=FALSE, warning=FALSE}
source("R/common.R")
library(tidyverse)

cat("Metadata\n\n")
load("../iMAP_part1/RDataRDS/metadata_objects.RData", verbose = T)

cat("\nOtu and taxonomy tables\n\n")
load("RDataRDS/otutaxonomy_objects.RData", verbose = T)
```

### Creating R objects from `mothur` output {#mothur-composite}
```{r}
taxlevels <- c("kingdom", "phylum", "class", "order", "family", "genus" )

mo_metadata <-  mo_metadata %>% 
  select(sample_id, isolate) %>% 
  drop_na(isolate)

mo_otu_counts <- mo_otutable %>% 
  pivot_longer(-otu, names_to = "sample_id", values_to = "count") %>% 
  relocate(sample_id, .before = otu)

mo_taxonomy <- mo_taxonomy 
  
mo_composite <- inner_join(mo_metadata, mo_otu_counts, by = "sample_id") %>% 
  inner_join( ., mo_taxonomy, by = "otu") %>% 
  group_by(sample_id) %>% 
  mutate(rel_abund = count/sum(count)) %>% 
  ungroup() %>% 
  relocate(count, .before = rel_abund) 

cat("\nColumn names of the composite file\n")
colnames(mo_composite)

cat("\nDoes the relative abundance adds to 1? Must!\n")
mo_composite %>% 
  group_by(sample_id) %>% 
  summarise(total_re_abund = sum(rel_abund))

mo_Rjoined_object <- mo_composite %>% 
  pivot_longer(cols = c(all_of(taxlevels), "otu"), names_to = "level", values_to = "taxon") %>% 
  relocate(rel_abund, .after = taxon)


saveRDS(mo_composite, "RDataRDS/mo_composite.rds")
save(mo_Rjoined_object, file = "RDataRDS/Rjoined_objects.RData")
save(mo_composite, mo_Rjoined_object, file = "RDataRDS/saved_objects.RData")
```



### Creating R objects from `qiime2` output {#qiime2-composite}
```{r}
taxlevels <- c("kingdom", "phylum", "class", "order", "family", "genus" )

metadata <- q2_metadata %>% 
  select(sample_id = "sample-id", isolate) %>% 
  drop_na(isolate)

otu_counts <- q2_otutable %>% 
  pivot_longer(-feature, names_to = "sample_id", values_to = "count") %>% 
  relocate(sample_id, .before = feature)

taxonomy <- q2_taxonomy 
  
q2_composite <- inner_join(metadata, otu_counts, by = "sample_id") %>% 
  inner_join( ., taxonomy, by = "feature") %>% 
  group_by(sample_id) %>% 
  mutate(rel_abund = count/sum(count)) %>% 
  ungroup() %>% 
  relocate(count, .before = rel_abund)

cat("\nColumn names of the composite file\n")
colnames(q2_composite)

cat("\nDoes the relative abundance adds to 1? Must!\n")
q2_composite %>% 
  group_by(sample_id) %>% 
  summarise(total_re_abund = sum(rel_abund))

q2_Rjoined_object <- q2_composite %>% 
  pivot_longer(cols = c(all_of(taxlevels), "feature"), names_to = "level", values_to = "taxon") %>% 
  relocate(c(count, rel_abund), .after = taxon)

saveRDS(q2_composite, "RDataRDS/q2_composite.rds")
resave(q2_Rjoined_object, file = "RDataRDS/Rjoined_objects.RData")
resave(q2_composite, q2_Rjoined_object, file = "RDataRDS/saved_objects.RData")
```


### Creating R objects from `metaphlan` output  {#metaphlan-composite}
- Here we create a composite object using default relative abundance output.
```{r}
taxlevels <- c("kingdom", "phylum", "class", "order", "family", "genus" )

metadata <- m3_metadata %>% 
  select(sample_id, platform) %>% 
  drop_na(platform)

m3_rel_otutable <- m3_otutable %>% 
  pivot_longer(-otu, names_to = "sample_id", values_to = "rel_abund") %>% 
  relocate(sample_id, .before = otu)

taxonomy <- m3_taxonomy

m3_rel_composite <- inner_join(metadata, m3_rel_otutable, by = "sample_id") %>% 
  inner_join( ., m3_taxonomy, by = "otu")

cat("\nColumn names of the composite file\n")
colnames(m3_rel_composite)

cat("\nDoes the relative abundance adds to 1? Must!\n")
m3_rel_composite %>% 
  group_by(sample_id) %>% 
  summarise(total_re_abund = round(sum(rel_abund)))

m3_rel_Rjoined_object <- m3_rel_composite %>% 
  pivot_longer(cols = c(all_of(taxlevels), "otu"), names_to = "level", values_to = "taxon") %>% 
  relocate(rel_abund, .after = taxon)


saveRDS(m3_rel_composite, "RDataRDS/m3_rel_composite.rds")
resave(m3_rel_Rjoined_object, file = "RDataRDS/Rjoined_objects.RData")
resave(m3_rel_composite, m3_rel_Rjoined_object, file = "RDataRDS/saved_objects.RData")
```

### Creating R objects from `metaphlan` estimated read counts {#metaphlan-count-composite}
- Here we create a composite object containing abundance based on the estimated mapped reads.

```{r}
taxlevels <- c("kingdom", "phylum", "class", "order", "family", "genus" )

metadata <- m3_metadata %>% 
  select(sample_id, platform) %>% 
  drop_na(platform)

m3_read_based_otutable <-  m3_read_counts_otutable %>% 
  pivot_longer(-otu, names_to = "sample_id", values_to = "count") %>% 
  relocate(sample_id, .before = otu) %>% 
  group_by(sample_id) %>% 
  mutate(rel_abund = count/sum(count)) %>% 
  ungroup() %>% 
  relocate(count, .before = rel_abund) 

taxonomy <- m3_taxonomy

m3_read_count_composite <- inner_join(metadata, m3_read_based_otutable, by = "sample_id") %>% 
  inner_join( ., m3_taxonomy, by = "otu")

cat("\nColumn names of the composite file\n")
colnames(m3_read_count_composite)

cat("\nDoes the relative abundance adds to 1? Must!\n")
m3_read_count_composite %>% 
  group_by(sample_id) %>% 
  summarise(total_re_abund = round(sum(rel_abund)))

m3_read_count_Rjoined_object <- m3_read_count_composite %>% 
  pivot_longer(cols = c(all_of(taxlevels), "otu"), names_to = "level", values_to = "taxon") %>% 
  relocate(rel_abund, .after = taxon)

saveRDS(m3_read_count_composite, "RDataRDS/m3_read_count_composite.rds")
resave(m3_read_count_Rjoined_object, file = "RDataRDS/Rjoined_objects.RData")
resave(m3_read_count_Rjoined_object, file = "RDataRDS/saved_objects.RData")
```

### Review saved R joined objects
```{r}
load("RDataRDS/Rjoined_objects.RData", verbose = T)
```

## Creating joined pyloseq objects {#phyloseq-objects}

### Prerequisite
- Tidy metadata created in [PART 1](https://github.com/tmbuza/iMAP-part1/).
- Tidy microbial abundance table created in [PART 2](https://github.com/tmbuza/iMAP-part2/).
- Tidy taxonomy table created in [PART 2](https://github.com/tmbuza/iMAP-part2/).

### Load libraries and data
```{r message=FALSE, warning=FALSE}
source("R/common.R")
library(tidyverse)

cat("Metadata\n\n")
load("../iMAP_part1/RDataRDS/metadata_objects.RData", verbose = T)

cat("\nOtu and taxonomy tables\n\n")
load("RDataRDS/otutaxonomy_objects.RData", verbose = T)
```

### Using `mothur` output {#mothur-physeq}
```{r}
ps_mo_metadata <- mo_metadata %>% tibble::column_to_rownames("sample_id") %>% 
  sample_data(ps_mo_metadata)

ps_mo_otutable <- mo_otutable %>% tibble::column_to_rownames("otu") %>% 
  otu_table(ps_mo_otutable, taxa_are_rows = TRUE)

ps_mo_taxonomy <- as.matrix(mo_taxonomy %>% tibble::column_to_rownames("otu")) %>% 
  tax_table(ps_mo_taxonomy)

ps_mo <- merge_phyloseq(ps_mo_metadata, ps_mo_otutable, ps_mo_taxonomy)

ps_mo_taxa_tree = rtree(ntaxa(ps_mo), rooted=TRUE, tip.label=taxa_names(ps_mo))

ps_mo_raw <- phyloseq::merge_phyloseq(ps_mo, ps_mo_taxa_tree)
ps_mo_raw
ps_mo_rel <- phyloseq::transform_sample_counts(ps_mo_raw, function(x){x / sum(x)})

save(ps_mo_raw, ps_mo_rel,  file = "RDataRDS/phyloseq_objects.RData")
```

Summarizing `mothur` phyloseq object
```{r message=FALSE, warning=FALSE}
microbiome::summarize_phyloseq(ps_mo_raw)
```

### Using `qiime2` output {#qiime2-physeq}
```{r}
ps_q2_metadata <- q2_metadata %>% 
  rename(sample_id = "sample-id") %>% 
  tibble::column_to_rownames("sample_id") %>% 
  sample_data(ps_q2_metadata)

ps_q2_otutable <- q2_otutable %>% tibble::column_to_rownames("feature") %>% 
  otu_table(ps_q2_otutable, taxa_are_rows = TRUE)

ps_q2_taxonomy <- as.matrix(q2_taxonomy %>% tibble::column_to_rownames("feature")) %>% 
  tax_table(ps_q2_taxonomy)

ps_q2 <- merge_phyloseq(ps_q2_metadata, ps_q2_otutable, ps_q2_taxonomy)

ps_q2_taxa_tree = rtree(ntaxa(ps_q2), rooted=TRUE, tip.label=taxa_names(ps_q2))

ps_q2_raw <- phyloseq::merge_phyloseq(ps_q2, ps_q2_taxa_tree)
ps_q2_raw

ps_q2_rel <- phyloseq::transform_sample_counts(ps_q2_raw, function(x){x / sum(x)})

resave(ps_q2_raw, ps_q2_rel,  file = "RDataRDS/phyloseq_objects.RData")
```

Summarizing `qiime2` phyloseq object
```{r message=FALSE, warning=FALSE}
microbiome::summarize_phyloseq(ps_q2_raw)
```

### Using `metaphlan` output {#metaphlan-physeq}
```{r}
ps_m3_metadata <- m3_metadata %>% tibble::column_to_rownames("sample_id") %>% 
  sample_data(ps_m3_metadata)

ps_m3_otutable <- m3_read_counts_otutable %>% tibble::column_to_rownames("otu") %>% 
  otu_table(ps_m3_otutable, taxa_are_rows = TRUE)

ps_m3_taxonomy <- as.matrix(m3_read_counts_taxonomy %>% tibble::column_to_rownames("otu")) %>% 
  tax_table(ps_m3_taxonomy)

ps_m3 <- merge_phyloseq(ps_m3_metadata, ps_m3_otutable, ps_m3_taxonomy)

ps_m3_taxa_tree = rtree(ntaxa(ps_m3), rooted=TRUE, tip.label=taxa_names(ps_m3))

ps_m3_raw <- phyloseq::merge_phyloseq(ps_m3, ps_m3_taxa_tree)
ps_m3_raw

ps_m3_rel <- phyloseq::transform_sample_counts(ps_m3_raw, function(x){x / sum(x)})

resave(ps_m3_raw, ps_m3_rel,  file = "RDataRDS/phyloseq_objects.RData")
```


### Review saved phyloseq objects
```{r}
load("RDataRDS/phyloseq_objects.RData", verbose = T)
```
