# Data Tranformation {#data-transformation}

## Quick glimpse
Data transformation is intended to converting the values into ready-to-use matrices Here we will explore different way that are used to transform raw or relative abundances.

## Prerequisite
- Saved R objects
- Saved phyloseq objects

> Either of the R or phyloseq objects can be used as input data.

## Import libraries and saved objects
```{r}
set.seed(110912)
source("R/common.R")
library(tidyverse, suppressPackageStartupMessages())

cat("\nPhyloseq objects\n\n")
load("RDataRDS/phyloseq_objects.RData", verbose = T)
```
```{r}
library(phyloseq)
library(microbiome)

ps_raw <- ps_mo_raw
ps_rel <- ps_mo_rel
```

## Raw Abundance
```{r}
cat("\n\n")
otu_table(ps_raw)[1:5, 1:3]

```



##  No Transformation
> Similar to raw abundance

```{r}
(ps_identity <- microbiome::transform(ps_raw, 'identity'))

cat("\n\n")
otu_table(ps_identity)[1:5, 1:3]

```

## Relative abundance
```{r}
(ps_rel = phyloseq::transform_sample_counts(ps_raw, function(x){x / sum(x)}))

cat("\n\n")
otu_table(ps_rel)[1:5, 1:3]

```

## Arc sine (asin) transformation
- Typically used when dealing with proportional and percentages. 
- Proportionals range from 0 to 1
- Percentages range from 0 to 100
- The Metaphlan3 relative abundances are in percentages! That means the column totals in 100.

```{r}
x = otu_table(ps_rel)
y <- x/max(x)

ps_asin <- round(asin(sqrt(y)), 6)

ps_asin <- as.matrix(ps_asin)
ps_asin[1:5, 1:4]
```

## Compositional Version
> Similar to relative abundance

```{r}
(ps_compositional <- microbiome::transform(ps_raw, 'compositional'))

cat("\n\n")
otu_table(ps_compositional)[1:5, 1:3]
```

## Z-transform for OTUs

```{r message=FALSE, warning=FALSE}
(ps_z_otu <- microbiome::transform(ps_raw, 'Z', 'OTU'))

cat("\n\n")
otu_table(ps_z_otu)[1:5, 1:3]
```

## Z-transform for Samples

```{r message=FALSE, warning=FALSE}
(ps_z_sample <- microbiome::transform(ps_raw, 'Z', 'sample'))

cat("\n\n")
otu_table(ps_z_sample)[1:5, 1:3]
```

## Log10 Transform

```{r message=FALSE, warning=FALSE}
(ps_log10 <- microbiome::transform(ps_raw, 'log10'))

cat("\n\n")
otu_table(ps_log10)[1:5, 1:3]
```

## Log10p Transform

```{r message=FALSE, warning=FALSE}
(ps_log10p <- microbiome::transform(ps_raw, 'log10p'))

cat("\n\n")
otu_table(ps_log10p)[1:5, 1:3]
```

## CLR Transform

- Note that small pseudocount is added if data contains zeroes
```{r message=FALSE, warning=FALSE}
(ps_clr <- microbiome::transform(ps_raw, 'clr'))

cat("\n\n")
otu_table(ps_clr)[1:5, 1:3]

```


## Shift the baseline

```{r message=FALSE, warning=FALSE}
(ps_shift <- microbiome::transform(ps_raw, 'shift', shift=1))

cat("\n\n")
otu_table(ps_shift)[1:5, 1:3]

```

## Data Scaling

```{r message=FALSE, warning=FALSE}
(ps_scale <- microbiome::transform(ps_raw, 'scale', scale=1))

cat("\n\n")
otu_table(ps_scale)[1:5, 1:3]

```

## Transformed Objects
```{r}
save(
  ps_asin, 
  ps_identity,
  ps_compositional, 
  ps_z_otu, 
  ps_z_sample, 
  ps_log10, 
  ps_log10p, 
  ps_clr, 
  ps_shift, 
  ps_scale, 
  file = "RDataRDS/transformed_data_.RData")
```

## Current saved objects
```{r include=T}
resave(
  ps_asin, 
  ps_identity,
  ps_compositional, 
  ps_z_otu, 
  ps_z_sample, 
  ps_log10, 
  ps_log10p, 
  ps_clr, 
  ps_shift, 
  ps_scale,
  file = "RDataRDS/saved_objects.RData")
```