# Dimensionality Reduction {#data-reduction}

## Quick glimpse
Dimensionality reduction is intended to reduce the the dimension of the variables. Common methods used in microbiome data include:

- Linear methods
- Non-linear methods

## Prerequisite
- Saved R objects
- Saved phyloseq objects

> Either of the R or phyloseq objects can be used as input data.

## Import libraries and saved objects
```{r}
set.seed(110912)
source("R/common.R")
library(tidyverse, suppressPackageStartupMessages())

cat("\nPhyloseq objects\n\n")
load("RDataRDS/phyloseq_objects.RData", verbose = T)
```
```{r}
library(phyloseq)
library(microbiome)

ps_raw <- ps_mo_raw
ps_rel <- ps_mo_rel
```

```{block, type="tmbinfo", echo=T}
## Linear Dimensionality Reduction Methods
The most common and well known dimensionality reduction methods are the ones that apply linear transformations, like

### PCA (Principal Component Analysis)
Popularly used for dimensionality reduction in continuous data, PCA rotates and projects data along the direction of increasing variance. The features with the maximum variance are the principal components.

### Factor Analysis
A technique that is used to reduce a large number of variables into fewer numbers of factors. The values of observed data are expressed as functions of a number of possible causes in order to find which are the most important. The observations are assumed to be caused by a linear transformation of lower dimensional latent factors and added Gaussian noise.

### LDA (Linear Discriminant Analysis)
Projects data in a way that the class separability is maximised. Examples from same class are put closely together by the projection. Examples from different classes are placed far apart by the projection

Source: https://towardsdatascience.com/dimensionality-reduction-for-machine-learning-80a46c2ebb7e
```


## Reduced Objects
```{r}
save(

  file = "RDataRDS/reduced_data_.RData")
```

## Current saved objects
```{r include=T}
resave(

  file = "RDataRDS/saved_objects.RData")
```